---
title: ""
format: html

execute:
  echo: true
  warning: false
  message: false
---


```{r, echo = F}

library(kableExtra)


```



<div id="menu-toggle" onclick="toggleSidebar()">☰ Menu</div>

<div class="sidebar">
  <ul>
    <li><a href="index.qmd">Home</a></li>
    <li><a href="data.qmd">Data</a></li>
    <li><a href="methodology.qmd">Methodology</a></li>
    <li><a href="conclusions.qmd">Conclusions</a></li>
  </ul>
</div>

<div class="main-content">
  <!-- Your page content goes here -->

## Methodology {#methodology}

:::{.methodology-flow}

:::{.flow-step}
Data Preprocessing???
:::

:::{.flow-step}
Train/Val/Test Split????
:::

:::{.flow-step}
Neural Network Design???
:::

:::{.flow-step}
Model Training????
:::

:::{.flow-step}
Evaluation????
:::

:::

:::{.feature-grid}

:::{.card}
### 🔧 Data Preprocessing
Fore example: "Comprehensive data cleaning, handling missing values, outlier detection, and feature scaling to prepare the dataset for neural network training."

:::

:::{.card}
### 🧠 Neural Network Architecture
For example:"Implementation of deep neural networks using TensorFlow/Keras through R's reticulate package, with multiple hidden layers and dropout regularization".
:::

:::

```{r fake-split-summary, echo=FALSE}
library(kableExtra)
# Create fake data split summary table
split_summary <- data.frame(
  Dataset = c("Training", "Validation", "Test", "Total"),
  Samples = c(6403, 2134, 2134, 10671),
  Percentage = c("60%", "20%", "20%", "100%"),
  Class_Distribution = c("Balanced", "Balanced", "Balanced", "Original"))

kable(split_summary, 
      caption = "Data Split Summary") %>%
  kable_styling()
```

:::{.card}
### 📊 Model Evaluation Strategy
We used....

- **Accuracy:** .....
- ad more here

</div>

<script>
  function toggleSidebar() {
    const sidebar = document.querySelector('.sidebar');
    if (sidebar.classList.contains('open')) {
      sidebar.classList.remove('open');
    } else {
      sidebar.classList.add('open');
    }
  }
</script>