---
title: "EDA"
author: "Hope Hennessy"
date: "2025-09-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(corrplot)
library(lubridate)
```


* Alt (Altitude): Elevation of observation point above sea level (meters). 
* Aspect: Compass direction the slope faces. Range: 0-360° (0=North, 90=East, 180=South, 270=West). Note: North-facing slopes stay colder, south-facing get more sun
* Incline: Steepness/gradient of slope (degrees). Plausible Range: 0-60° (most avalanches occur 30-45°). Purpose: Critical for avalanche formation (too flat=no slide, too steep=continuous sluffing)
* Air.Temp (Air Temperature): Temperature at observation point(degrees celsius)
* Wind.Dir (Wind Direction): Direction wind is blowing FROM (degrees, 0-360). Purpose: Determines snow deposition patterns (lee slopes get loaded)
* Wind.Speed (m/s). Wind transports snow, creates dangerous loading
* Cloud (Cloud Cover): Sky coverage by clouds. Range: 0-100%
* Precip.Code (Precipitation Code): Type and intensity of precipitation. Categorical
* Drift (Snow Drift): Active snow transport by wind. Binary (0 = No, 1 = Yes)
* Rain.at.900: Rainfall at 900m elevation. Binary (0 = No, 1 = Yes). Purpose: Rain on snow = very dangerous 
* Summit.Air.Temp (degrees Celsius)
* Summit.Wind.Dir
* Summit.Wind.Speed (m/s)
* Total.Snow.Depth: Total height of snowpack (cm)
* Foot.Pen (Foot Penetration): How deep foot sinks into snow when walking (cm). Indicates snow consolidation/stability. Soft snow (high penetration) = unstable
* Ski.Pen (Ski Penetration): How deep skis sink into snow (cm)
* Max.Temp.Grad (Maximum Temperature Gradient): Largest temperature difference within snowpack. Large gradients can cause weak layers in the snow and increase avalanche risk.
* Max.Hardness.Grad (Maximum Hardness Gradient): Largest difference in snow hardness between layers
* No.Settle (New Snow Settlement): Amount of settlement/compaction of new snow
* Snow.Index: Composite snow stability index
* Insolation: Solar radiation/sun exposure. Sun warming affects snow stability
* Crystals: Crystal type/structure in snowpack
* Wetness: Moisture content of snow
* AV.Cat (Avalanche Category)
* Snow.Temp (Snow Temperature). Plausible Range: -20°C to 0°C



# Important Notes

```{r}
data <- read.csv("~/University/MSc Data Science/STA5073Z_DS Industry/DS-Industry/scotland_avalanche_forecasts_2009_2025.csv", header = TRUE)
```

```{r}
head(data)
dim(data)
str(data)
table(data$FAH) # target variable levels
summary(data)
```


# Cleaning

```{r}
# Only keep rows where the target variable FAH is not missing (-109 rows)
data <- data %>% filter(!is.na(FAH) & FAH != "")

# Convert categorical target variable to numeric 
data$FAH <- as.integer(factor(x = data$FAH, levels = c("Low", "Moderate", "Considerable -", "Considerable +", "High"))) - 1
data$Precip.Code <- as.integer(factor(data$Precip.Code)) - 1

# Create a year variable from date
# Don't have any data for July, Aug and Sept (summer months) - don't know if a month variable will be important?
data <- data %>% mutate(Year = year(Date))

# Remove unnecessary columns
data <- data %>% select(-OSgrid, -OAH, -Obs)
```

# Missing data summary

```{r}
numeric_vars <- data %>% select(where(is.numeric))

missing_summary <- numeric_vars %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percent = round(Missing_Count / nrow(data) * 100, 1)) %>%
  arrange(desc(Missing_Count))
missing_summary

```

# Visual Outliers

```{r}
# Simplified grouped boxplots
plot_group_boxplots <- function(vars, df, title, color) {
  plots <- lapply(vars, function(v) {
    ggplot(df, aes(y = .data[[v]])) +
      geom_boxplot(fill = color, alpha = 0.6) +
      labs(title = v, y = v) +
      theme_minimal()
  })
  do.call(grid.arrange, c(plots, ncol = 3, top = title))
}

# Plotting - these groups are random...Just exploring...
plot_group_boxplots(c("Alt", "Aspect", "Incline"), numeric_vars, "Terrain Variables", "steelblue")
plot_group_boxplots(c("Air.Temp", "Wind.Speed", "Cloud"), numeric_vars, "Weather Variables", "orange")
plot_group_boxplots(c("Total.Snow.Depth", "Foot.Pen", "Max.Temp.Grad"), numeric_vars, "Snow Pack Variables", "purple")
plot_group_boxplots(c("Max.Temp.Grad", "No.Settle", "Snow.Index"), numeric_vars, "Variables", "steelblue")
plot_group_boxplots(c("Insolation", "Snow.Temp"), numeric_vars, "Variables", "steelblue")

```



* The highest mountain in Scotland is Ben Nevis, which has an altitude of 1,345 meters above sea level. Our data has to impossible outliers (77044 & 244859) = remove!

* "It is worthy of note that the highest wind speed recorded at the summit weather station on Cairngorm at the time of writing was an eye-watering 176 mph!" (78.7 m/s) https://www.mountaineering.scot/safety-and-skills/essential-skills/weather-conditions/wind-speeds

* In March 1986, on the summit of the Cairngorms in the Highlands of Scotland, a wind gust of 173mph (278km/h) was recorded.: https://www.bbc.com/weather/articles/cvglvnyxpx0o

* Don't quite know how to interpret Snow.Index. 737 missing. 7677 zero values.

```{r}
# Cleaning dataset from errors 
data_cleaned <- data %>%
  mutate(
    # Altitude (0 to 1345): removing Alt values 77044 and 244859	
    Alt = ifelse(Alt < 0 | Alt > 1345, NA_real_, Alt),
    
    # Aspect (0 to 360)
    Aspect = ifelse(Aspect < 0 | Aspect > 360, NA_real_, Aspect),
    
    # Incline (0 to 90)
    Incline = ifelse(Incline < 0 | Incline > 90, NA_real_, Incline),
    
    # Wind Speed - look above to reference of highest wind speed in Scotland
    Wind.Speed = ifelse(Wind.Speed < 0 | Wind.Speed > 79, NA_real_, Wind.Speed),
    
    # Cloud Cover (0 to 100) (there is just 3 wierd values: -1, -1, 199)
    Cloud = ifelse(Cloud < 0 | Cloud > 100, NA_real_, Cloud),
    
    # Clean Total Snow Depth (0 to 1500). Removing values: 2000, 2000, 2500, 3000, 2010, 2080
    Total.Snow.Depth = ifelse(Total.Snow.Depth < 0 | Total.Snow.Depth > 1500, NA_real_, Total.Snow.Depth),
    
    # Summit Wind Speed (m/s) (same threshold as Wind Speed): removed quite a few values
    Summit.Wind.Speed = ifelse(Summit.Wind.Speed < 0 | Summit.Wind.Speed > 79, NA_real_, Summit.Wind.Speed),
    
    # Foot Penetration (depth in cm) - THIS NEEDS SOME RESEARCH
    Foot.Pen = ifelse(Foot.Pen < 0 | Foot.Pen > 200, NA_real_, Foot.Pen),
    
    # Wind direction
    Wind.Dir = ifelse(Wind.Dir < 0 | Wind.Dir > 360, NA_real_, Wind.Dir),
    
    # Summit wind direction. Removed: 370, 385, 390, 2213
    # 2213 is obviously wrong but maybe 370 means 10, 385 means 25 and 390 means 30????
    
    Summit.Wind.Dir = ifelse(Summit.Wind.Dir > 360 & Summit.Wind.Dir <= 720, 
                              Summit.Wind.Dir - 360, Summit.Wind.Dir), # Should i add this line in???
    
    Summit.Wind.Dir = ifelse(Summit.Wind.Dir < 0 | Summit.Wind.Dir > 360, NA_real_, Summit.Wind.Dir),
    
    # New Snow Settlement (cant be negative values). There are three large outliers (> 600)
    No.Settle = ifelse(No.Settle < 0, NA_real_, No.Settle),
    
    # Insolation
    # Assuming it is Daily Solar Radiation (MJ/m²/day): remove values > 30 (outlier values 106 & 208 removed)
    Insolation = ifelse(Insolation < 0 | Insolation > 30, NA_real_, Insolation))

```


Aspect & wind direction are circular variables, where 0° and 360° represent the same direction, which can confuse models that treat it as linear. By converting Aspect to sine and cosine, you preserve its circular nature and allow the model to capture directional relationships correctly.

```{r}
# Should we do this??
data_cleaned <- data_cleaned %>% 
  mutate(Aspect_sin = sin(Aspect * pi / 180), Aspect_cos = cos(Aspect * pi / 180)) %>% 
  mutate(WindDir_sin = sin(Wind.Dir * pi / 180), WindDir_cos = cos(Wind.Dir * pi / 180)) %>%
  mutate(SummitWindDir_sin = sin(Summit.Wind.Dir * pi / 180), SummitWindDir_cos = cos(Summit.Wind.Dir * pi / 180)) %>%
  select(-Wind.Dir, -Aspect, -Summit.Wind.Dir)

```


## Investigating Ski Pen variable

* Ski.Pen: 7485 rows of values 0. 2375 rows of NA values.
* The thing is, you'd expect large foot.pen values to correspond with sk.pin values greater than 0, but im not seeing that? Ski.pen seems very sus.
* Ski.pen = 0 for very large values of foot.pen...Cnt be right
* It doesn't seem worth keeping this variable in the analysis???

```{r}
nrow(data[data$Ski.Pen == 0 & !is.na(data$Ski.Pen), ])
data[data$Ski.Pen == 0 & data$Foot.Pen > 50 & !is.na(data$Ski.Pen) & !is.na(data$Foot.Pen), ]

# Removing rows with NA in Ski.Pen or Foot.Pen
clean_data <- data[!is.na(data$Ski.Pen) & !is.na(data$Foot.Pen), ]

plot(clean_data$Ski.Pen, clean_data$Foot.Pen, xlab = "Ski Penetration", 
     ylab = "Foot Penetration",pch = 19)
```


## Investigating Max.Temp.Grad

* 4635 rows of values = 0
* 642 NA.
* Must do more research here

```{r}
summary(data$Max.Temp.Grad)


data[data$Max.Temp.Grad > 40 & !is.na(data$Max.Temp.Grad), ]

nrow(data[data$Max.Temp.Grad == 0 & !is.na(data$Max.Temp.Grad), ])

summary(data$Max.Temp.Grad, na.rm = TRUE)
na.omit(data$Max.Temp.Grad)


```

## Investigating Crystals & Wetness

How to deal with this variable?

```{r}
table(data$Crystals)
# -1: Missing ???
# 0: Very fine grains, fresh snow, or rounded grains (<0.5mm)??
# 1: Small grains (1mm)
# 2: Medium grains (2mm)
# 5: Large grains (5mm) 
# 8: Very large grains
# 10: Extreme grain size 
# 11 & 15: Exceptionally large crystals 


table(data$Wetness) # -1, 0, 1, 2,3, 4, 6, 10 ????

```

## Investigating AV.Cat (likely stands for avalanche category)

This just seems sus... 
It would be nice to leave this variable out...?

```{r}
table(data$AV.Cat, useNA = "ifany")

summary(data$AV.Cat)

unique(data$AV.Cat)

# Check relationship with FAH
table(data$AV.Cat, data$FAH)

# Missing Vlaues
sum(is.na(data$AV.Cat) | data$AV.Cat == "NA" | data$AV.Cat == -1)
```

## Investigating Snow.Temp

Something is very wrong here. Snow melts > 0 degrees Celsius. What to do about this?

```{r}
# What temp values are > 0 degrees celcius - 278 rows!!!
data[data$Snow.Temp > 0 & !is.na(data$Snow.Temp), ]

summary(data$Snow.Temp, na.rm = TRUE)
na.omit(data$Snow.Temp)


# Check the distribution of these errors
summary(data$Snow.Temp[data$Snow.Temp > 0 & !is.na(data$Snow.Temp)])

# Are they clustered around certain values?
table(data$Snow.Temp[data$Snow.Temp > 0 & !is.na(data$Snow.Temp)])

# These rows compared to Air.Temp...Somethings very wrong
data %>%
  filter(Snow.Temp > 0) %>%
  select(Date, Snow.Temp, Air.Temp) 
```






# Correlation Analysis

```{r}
numeric_vars <- data_cleaned %>% select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")

corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         order = "hclust",
         tl.cex = 0.7,        
         tl.col = "black",     
         tl.srt = 45, 
         title = "Correlation Matrix - Scottish Avalanche Data",
         mar = c(0,0,2,0))

```









