---
title: ""  # Empty title since we use custom hero
format: 
  html:
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(kableExtra)
library(ggplot2)

# Set theme for all plots
theme_set(theme_minimal() + 
          theme(plot.background = element_rect(fill = "white", color = NA),
                panel.background = element_rect(fill = "white", color = NA)))

# Set seed for reproducible fake data
set.seed(123)
```

:::{.hero}
# Avalanche Hazard Forecasting using  {.text-center}

:::{.hero-subtitle}
Predicting forecasted avalanche hazard levels from 15 years of Scottish data using a neural network.
:::
:::

## Overview {#overview}

:::{.feature-grid}

:::{.card}
### 🎯 Objective
Develop and evaluate a neural network model......
:::

:::{.card}
### 📊 Dataset  
15 years of Scottish avalanche data. 10,671 records with 34 variables.....
:::

:::{.card}
### 🧠 Approach
How did we go about this?
:::

:::

```{r data-overview, eval = FALSE, echo = TRUE}
# Data loading and initial exploration
# This code will be executed once actual data is available

# Load the Scottish avalanche dataset
scottish_avalanche_data <- read.csv("scottish_avalanche_data.csv")

# Display basic dataset information
glimpse(scottish_avalanche_data)

# Check for missing values
missing_summary <- scottish_avalanche_data %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  arrange(desc(Missing_Count))

print(missing_summary)

# Basic statistical summary
summary(scottish_avalanche_data)
```

```{r fake-data-display, echo = FALSE}

# Create fake summary table for display purposes
fake_summary <- data.frame(
  Variable = c("Date", "Region", "Forecasted_Hazard", "Temperature_Min", "Temperature_Max", 
               "Wind_Speed", "Precipitation", "Snow_Depth", "Elevation", "Aspect"),
  Type = c("Date", "Factor", "Integer", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Integer", "Factor"),
  Missing_Count = c(0, 0, 15, 234, 189, 123, 567, 89, 45, 12),
  Description = c("Observation date", "Forecasting region (6 levels)", "Hazard level 1-5", 
                 "Daily minimum temperature (°C)", "Daily maximum temperature (°C)",
                 "Wind speed (m/s)", "Daily precipitation (mm)", "Snow depth (cm)",
                 "Elevation (m)", "Slope aspect (8 directions)")
)

kable(fake_summary, 
      caption = "Dataset Variables Overview (Sample of 34 total variables)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Dataset & Target Variable {#data}

### Avalanche Hazard Scale

:::{.hazard-scale}

:::{.hazard-level .level-1}
**Level 1**  
This is 1 (?)
:::

:::{.hazard-level .level-2}
**Level 2**  
This is 2 (?)
:::

:::{.hazard-level .level-3}
**Level 3**  
This is 3 (?)
:::

:::{.hazard-level .level-4}
**Level 4**  
This is 4 (?)
:::

:::{.hazard-level .level-5}
**Level 5**  
This is 5 (?)
:::

:::

:::{.feature-grid}

:::{.card}
### 🌍 Location/Topography Features
- Forecasting region (......)
- Add
- The
- features
- here
:::

:::{.card}
### 🌦️ Weather Conditions
- Temperature measurements...
- Add
the 
features
here

:::

:::{.card}
### ❄️ Snow Pack Measurements  
- Snow....
add 
the
features
here
:::

:::

```{r correlation-analysis, eval = FALSE, echo =TRUE}
# Correlation analysis of environmental variables
# This code will analyze relationships between variables once data is loaded

# Select numeric variables for correlation analysis
numeric_vars <- scottish_avalanche_data %>%
  select(forecasted_hazard, temperature_min, temperature_max, 
         wind_speed, precipitation, snow_depth, elevation,
         humidity, pressure, new_snow_24h, wind_direction_degrees)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Create correlation plot
library(corrplot)
corrplot(cor_matrix, 
         method = "color",
         type = "upper", 
         order = "hclust",
         tl.cex = 0.8,
         tl.col = "black",
         tl.srt = 45,
         title = "Environmental Variables Correlation Matrix")

# Identify highly correlated features (>0.8 or <-0.8)
high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
print("Highly correlated variable pairs:")
print(high_cor)
```

```{r fake-correlation-table, echo = FALSE}
# Fake correlation data for display
fake_correlations <- data.frame(
  Variable_1 = c("Forecasted_Hazard", "Snow_Depth", "Temperature_Min", "Wind_Speed", "Precipitation"),
  Variable_2 = c("Snow_Depth", "New_Snow_24h", "Temperature_Max", "Wind_Direction", "Humidity"),
  Correlation = c(0.67, 0.82, 0.91, -0.34, 0.45),
  Interpretation = c("Strong positive", "Very strong positive", "Very strong positive", "Moderate negative", "Moderate positive")
)

kable(fake_correlations, 
      caption = "Key Variable Correlations (Sample Results)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Methodology {#methodology}

:::{.methodology-flow}

:::{.flow-step}
Data Preprocessing???
:::

:::{.flow-step}
Train/Val/Test Split????
:::

:::{.flow-step}
Neural Network Design???
:::

:::{.flow-step}
Model Training????
:::

:::{.flow-step}
Evaluation????
:::

:::

:::{.feature-grid}

:::{.card}
### 🔧 Data Preprocessing
Fore example: "Comprehensive data cleaning, handling missing values, outlier detection, and feature scaling to prepare the dataset for neural network training."

:::

:::{.card}
### 🧠 Neural Network Architecture
For example:"Implementation of deep neural networks using TensorFlow/Keras through R's reticulate package, with multiple hidden layers and dropout regularization".
:::

:::

```{r train-test-split, eval = FALSE, echo = TRUE}
# Data splitting strategy
library(caret)

# Create stratified train/validation/test split (60/20/20)
set.seed(123)

# First split: separate test set (20%)
train_val_index <- createDataPartition(scaled_data$forecasted_hazard, 
                                      p = 0.8, list = FALSE)
train_val_data <- scaled_data[train_val_index, ]
test_data <- scaled_data[-train_val_index, ]

# Second split: separate training and validation (60/20 of original)
train_index <- createDataPartition(train_val_data$forecasted_hazard, 
                                  p = 0.75, list = FALSE)  # 0.75 * 0.8 = 0.6
train_data <- train_val_data[train_index, ]
val_data <- train_val_data[-train_index, ]

# Prepare features and targets
X_train <- train_data %>% select(-forecasted_hazard) %>% as.matrix()
y_train <- train_data$forecasted_hazard - 1  # Convert to 0-4 for keras

X_val <- val_data %>% select(-forecasted_hazard) %>% as.matrix()
y_val <- val_data$forecasted_hazard - 1

X_test <- test_data %>% select(-forecasted_hazard) %>% as.matrix()
y_test <- test_data$forecasted_hazard - 1

# Print data split summary
cat("Training samples:", nrow(train_data), "\n")
cat("Validation samples:", nrow(val_data), "\n") 
cat("Test samples:", nrow(test_data), "\n")
```

```{r fake-split-summary, echo=FALSE}

# Create fake data split summary table
split_summary <- data.frame(
  Dataset = c("Training", "Validation", "Test", "Total"),
  Samples = c(6403, 2134, 2134, 10671),
  Percentage = c("60%", "20%", "20%", "100%"),
  Class_Distribution = c("Balanced", "Balanced", "Balanced", "Original"))

kable(split_summary, 
      caption = "Data Split Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

:::{.card}
### 📊 Model Evaluation Strategy
We used....

- **Accuracy:** .....
- ad more here

:::

## Results & Analysis {#results}

```{r training-history, eval = FALSE, echo = TRUE}
# Neural network training
# Set up callbacks for training monitoring
callbacks_list <- list(
  callback_early_stopping(monitor = "val_loss", patience = 10, restore_best_weights = TRUE),
  callback_reduce_lr_on_plateau(monitor = "val_loss", factor = 0.5, patience = 5),
  callback_model_checkpoint("best_avalanche_model.h5", monitor = "val_accuracy", 
                           save_best_only = TRUE))

# Train the model
history <- model %>% fit(
  X_train, y_train,
  epochs = 100,
  batch_size = 32,
  validation_data = list(X_val, y_val),
  callbacks = callbacks_list,
  verbose = 1)

# Plot training history
plot(history)

# Load best model and make predictions
best_model <- load_model_hdf5("best_avalanche_model.h5")
predictions <- best_model %>% predict(X_test)
predicted_classes <- apply(predictions, 1, which.max) - 1
```

```{r fake-performance-metrics, echo = FALSE}
# Create fake performance metrics table
performance_results <- data.frame(
  Metric = c("Overall Accuracy", "Weighted Precision", "Weighted Recall", "Weighted F1-Score", 
            "Macro F1-Score", "Cohen's Kappa", "Multi-class AUC", "Training Time"),
  Value = c("0.847", "0.851", "0.847", "0.845", "0.763", "0.789", "0.923", "47 minutes"),
  Standard_Error = c("±0.012", "±0.015", "±0.012", "±0.014", "±0.023", "±0.018", "±0.008", "N/A"))

kable(performance_results, 
      caption = "Neural Network Performance Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r fake-confusion-matrix, echo = FALSE}
library(dplyr)
library(tidyr)
library(kableExtra)

# Create fake confusion matrix data
confusion_data <- data.frame(
  Actual = rep(1:5, each = 5),
  Predicted = rep(1:5, times = 5),
  Count = c(
    387, 23, 8, 2, 0,    # Actual 1
    45, 512, 67, 12, 1,   # Actual 2  
    12, 89, 678, 89, 15,  # Actual 3
    3, 18, 76, 234, 34,   # Actual 4
    0, 2, 8, 23, 67        # Actual 5
  )
)

# Pivot to wide format
confusion_table <- confusion_data %>%
  pivot_wider(names_from = Predicted, values_from = Count,
              names_prefix = "Pred_") %>%
  select(-Actual)

# Add row names
rownames(confusion_table) <- paste("Actual", 1:5)

# Display confusion matrix
kable(confusion_table, 
      caption = "Confusion Matrix - Test Set Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c("Predicted Hazard Level" = ncol(confusion_table)))

```



### Class-Specific Performance Analysis

```{r}
#| label: fake-class-metrics
#| echo: false

# Create fake per-class performance metrics
class_metrics <- data.frame(
  Hazard_Level = c("Level 1 (Low)", "Level 2 (Moderate)", "Level 3 (Considerable)", 
                  "Level 4 (High)", "Level 5 (Extreme)"),
  Precision = c(0.891, 0.796, 0.821, 0.641, 0.456),
  Recall = c(0.921, 0.805, 0.769, 0.640, 0.582),
  F1_Score = c(0.906, 0.800, 0.794, 0.641, 0.512),
  Support = c(420, 637, 883, 365, 100),
  Notes = c("Excellent", "Very Good", "Good", "Acceptable", "Challenging due to low frequency")
)

kable(class_metrics, 
      caption = "Per-Class Performance Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(6, width = "3cm")
```

:::{.results-grid}

:::{.chart-placeholder}
**📈 Training & Validation Curves**  
*Learning curves showing model convergence over 67 epochs with early stopping*
:::

:::{.chart-placeholder}
**📊 ROC Curves by Class**  
*Multi-class ROC analysis with AUC values ranging from 0.89 to 0.96*
:::

:::

```{r}
#| label: feature-importance-analysis
#| eval: false
#| echo: true

# Feature importance analysis using SHAP values
library(shapr)

# Calculate SHAP values for feature importance
explainer <- shapr(X_train, model)
shap_values <- explain(X_test[1:100, ], explainer, approach = "empirical")

# Aggregate SHAP values for global feature importance
feature_importance <- data.frame(
  Feature = colnames(X_train),
  Importance = apply(abs(shap_values$dt), 2, mean)
) %>%
  arrange(desc(Importance)) %>%
  head(15)  # Top 15 features

# Alternative: Use permutation importance
library(vip)
vi_scores <- vi(model, method = "permute", train = cbind(X_train, y_train),
               target = "y_train", metric = "accuracy")
```

```{r}
#| label: fake-feature-importance
#| echo: false

# Create fake feature importance results
feature_importance_results <- data.frame(
  Rank = 1:15,
  Feature = c("Snow_Depth_Current", "New_Snow_24h", "Wind_Speed_Mean", "Temperature_Min",
             "Snow_Stability_Index", "Precipitation_48h", "Wind_Direction_Consistency",
             "Temperature_Gradient", "Elevation", "Aspect_Northness", "Humidity_Mean",
             "Cloud_Cover", "Temperature_Max", "Surface_Conditions", "Previous_Avalanche_Activity"),
  Importance_Score = c(0.234, 0.187, 0.156, 0.143, 0.134, 0.121, 0.108, 0.097,
                      0.089, 0.076, 0.071, 0.063, 0.058, 0.052, 0.047),
  Category = c("Snow", "Snow", "Weather", "Weather", "Snow", "Weather", "Weather",
              "Snow", "Terrain", "Terrain", "Weather", "Weather", "Weather", "Snow", "Historical")
)

kable(feature_importance_results, 
      caption = "Top 15 Most Important Features (SHAP Analysis)") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(1:3, bold = TRUE, background = "#f0f8ff")
```

## Cross-Validation & Model Robustness {#validation}

```{r}
#| label: cross-validation-setup
#| eval: false
#| echo: true

# Comprehensive cross-validation analysis
library(caret)

# Stratified K-Fold Cross-Validation (k=10)
set.seed(123)
cv_folds <- createFolds(scaled_data$forecasted_hazard, k = 10, list = TRUE, returnTrain = TRUE)

# Function to train and evaluate model on each fold
cv_results <- data.frame()

for(i in 1:10) {
  # Split data for current fold
  train_indices <- cv_folds[[i]]
  cv_train <- scaled_data[train_indices, ]
  cv_test <- scaled_data[-train_indices, ]
  
  # Prepare data for neural network
  X_cv_train <- cv_train %>% select(-forecasted_hazard) %>% as.matrix()
  y_cv_train <- cv_train$forecasted_hazard - 1
  X_cv_test <- cv_test %>% select(-forecasted_hazard) %>% as.matrix()
  y_cv_test <- cv_test$forecasted_hazard - 1
  
  # Train model (reduced epochs for CV)
  cv_model <- create_model()  # Function to create fresh model
  cv_history <- cv_model %>% fit(
    X_cv_train, y_cv_train,
    epochs = 50,
    batch_size = 32,
    validation_split = 0.2,
    verbose = 0
  )
  
  # Evaluate on test fold
  cv_predictions <- cv_model %>% predict(X_cv_test)
  cv_pred_classes <- apply(cv_predictions, 1, which.max) - 1
  
  # Calculate metrics
  fold_accuracy <- mean(cv_pred_classes == y_cv_test)
  fold_f1 <- F1_Score(y_cv_test, cv_pred_classes, average = "weighted")
  
  # Store results
  cv_results <- rbind(cv_results, data.frame(
    Fold = i,
    Accuracy = fold_accuracy,
    F1_Score = fold_f1
  ))
}

# Summary statistics
cv_summary <- cv_results %>%
  summarise(
    Mean_Accuracy = mean(Accuracy),
    SD_Accuracy = sd(Accuracy),
    Mean_F1 = mean(F1_Score),
    SD_F1 = sd(F1_Score)
  )
```

```{r}
#| label: fake-cv-results
#| echo: false

# Create fake cross-validation results
cv_fake_results <- data.frame(
  Fold = 1:10,
  Accuracy = c(0.856, 0.834, 0.851, 0.847, 0.839, 0.862, 0.841, 0.853, 0.844, 0.858),
  F1_Score = c(0.851, 0.829, 0.847, 0.843, 0.834, 0.857, 0.836, 0.849, 0.840, 0.853),
  Training_Time_Min = c(4.2, 4.1, 4.3, 4.0, 4.2, 4.4, 4.1, 4.3, 4.0, 4.2)
)

# Add summary row
cv_summary_row <- data.frame(
  Fold = "Mean ± SD",
  Accuracy = "0.847 ± 0.009",
  F1_Score = "0.843 ± 0.008", 
  Training_Time_Min = "4.2 ± 0.1"
)

# Combine and display
cv_display <- rbind(cv_fake_results, cv_summary_row)
cv_display[11, 1] <- "**Mean ± SD**"

kable(cv_display, 
      caption = "10-Fold Cross-Validation Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(11, bold = TRUE, background = "#f0f8ff")
```

## Regional & Seasonal Analysis {#regional}

```{r}
#| label: regional-performance
#| eval: false
#| echo: true

# Analyze model performance by region and season
# This analysis helps identify if the model generalizes across different contexts

# Performance by region
regional_analysis <- test_data %>%
  mutate(
    predicted_hazard = predicted_classes + 1,
    correct = (forecasted_hazard == predicted_hazard)
  ) %>%
  group_by(region) %>%
  summarise(
    n_observations = n(),
    accuracy = mean(correct),
    precision = precision_func(forecasted_hazard, predicted_hazard),
    recall = recall_func(forecasted_hazard, predicted_hazard),
    .groups = 'drop'
  )

# Seasonal analysis
seasonal_analysis <- test_data %>%
  mutate(
    predicted_hazard = predicted_classes + 1,
    correct = (forecasted_hazard == predicted_hazard),
    month = month(date),
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer", 
      month %in% c(9, 10, 11) ~ "Autumn"
    )
  ) %>%
  group_by(season) %>%
  summarise(
    n_observations = n(),
    accuracy = mean(correct),
    avg_actual_hazard = mean(forecasted_hazard),
    avg_predicted_hazard = mean(predicted_hazard),
    .groups = 'drop'
  )
```

```{r}
#| label: fake-regional-results
#| echo: false

# Fake regional performance data
regional_performance <- data.frame(
  Region = c("Lochaber", "Glen Coe", "Creag Meagaidh", "Southern Cairngorms", 
            "Northern Cairngorms", "Torridon"),
  Observations = c(389, 342, 356, 398, 367, 282),
  Accuracy = c(0.851, 0.834, 0.862, 0.841, 0.856, 0.829),
  Precision = c(0.847, 0.831, 0.859, 0.838, 0.853, 0.826),
  Recall = c(0.851, 0.834, 0.862, 0.841, 0.856, 0.829),
  F1_Score = c(0.849, 0.833, 0.860, 0.840, 0.855, 0.827)
)

kable(regional_performance, 
      caption = "Model Performance by Scottish Forecasting Region") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(c(3,4,5,6), width = "2cm")
```

```{r}
#| label: fake-seasonal-results
#| echo: false

# Fake seasonal performance data
seasonal_performance <- data.frame(
  Season = c("Winter (Dec-Feb)", "Spring (Mar-May)", "Summer (Jun-Aug)", "Autumn (Sep-Nov)"),
  Observations = c(892, 634, 278, 530),
  Accuracy = c(0.863, 0.821, 0.744, 0.835),
  Avg_Actual_Hazard = c(2.8, 2.1, 1.4, 2.3),
  Avg_Predicted_Hazard = c(2.7, 2.2, 1.6, 2.2),
  Notes = c("Peak season, best performance", "Good performance", 
           "Limited data, lower performance", "Moderate performance")
)

kable(seasonal_performance, 
      caption = "Model Performance by Season") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(6, width = "4cm")
```

## AI Tools Integration {#ai-tools}

:::{.feature-grid}

:::{.card}
### 🤖 ChatGPT & Claude Utilization (This text is example!!!!)
**Code Development:** Assisted with neural network architecture design, hyperparameter tuning strategies, and debugging complex TensorFlow/Keras implementations.

**Data Analysis:** Provided guidance on appropriate evaluation metrics for imbalanced multi-class problems and suggested advanced visualization techniques for model interpretation.

**Literature Review:** Helped identify relevant research papers on avalanche prediction and machine learning applications in glaciology and mountain safety.

**Documentation:** Supported in structuring the technical report, creating clear explanations of complex statistical concepts, and ensuring accessibility for different audiences.
:::

:::{.card}
### 📝 Critical Assessment & Limitations (This text is example!!!!)
**Strengths:**
- Rapid prototyping and code generation
- Comprehensive literature synthesis
- Error detection and debugging assistance
- Multiple approach suggestions and alternatives

**Limitations:**
- Required domain expertise validation for avalanche-specific considerations
- Manual verification of all statistical methods and model architectures
- Careful review needed for safety-critical recommendations
- Cannot replace human judgment in life-safety applications

**Best Practices Adopted:**
- Used AI as collaborative tool while maintaining critical thinking
- Cross-referenced all technical recommendations with peer-reviewed literature
- Implemented multiple validation approaches to ensure model reliability
- Maintained transparency about AI assistance throughout the research process
:::

:::

## Error Analysis & Model Interpretation {#interpretation}

```{r}
#| label: error-analysis
#| eval: false
#| echo: true

# Detailed error analysis to understand model failures
# This helps identify systematic biases and improvement opportunities

# Analyze misclassifications
error_analysis <- test_data %>%
  mutate(
    predicted_hazard = predicted_classes + 1,
    prediction_error = predicted_hazard - forecasted_hazard,
    error_type = case_when(
      prediction_error == 0 ~ "Correct",
      prediction_error > 0 ~ "Over-predicted",
      prediction_error < 0 ~ "Under-predicted"
    ),
    error_magnitude = abs(prediction_error)
  )

# Critical errors (off by 2+ levels)
critical_errors <- error_analysis %>%
  filter(error_magnitude >= 2) %>%
  select(date, region, forecasted_hazard, predicted_hazard, 
         snow_depth, temperature_min, wind_speed) %>%
  arrange(desc(error_magnitude))

# Error patterns by environmental conditions
error_patterns <- error_analysis %>%
  group_by(error_type) %>%
  summarise(
    count = n(),
    avg_snow_depth = mean(snow_depth, na.rm = TRUE),
    avg_temp_min = mean(temperature_min, na.rm = TRUE),
    avg_wind_speed = mean(wind_speed, na.rm = TRUE),
    .groups = 'drop'
  )

print("Critical prediction errors (≥2 levels off):")
print(critical_errors)
```

```{r}
#| label: fake-error-analysis
#| echo: false

# Create fake error analysis results
error_summary <- data.frame(
  Error_Type = c("Correct Predictions", "Over-predicted by 1", "Under-predicted by 1", 
                "Over-predicted by 2+", "Under-predicted by 2+"),
  Count = c(1806, 187, 231, 23, 87),
  Percentage = c("77.1%", "8.0%", "9.9%", "1.0%", "3.7%"),
  Avg_Snow_Depth = c(67.2, 45.3, 89.4, 23.1, 112.7),
  Avg_Wind_Speed = c(12.4, 18.7, 8.9, 24.3, 6.2),
  Common_Conditions = c("Typical conditions", "High wind, low snow", "Deep snow, calm",
                       "Extreme wind", "Very deep snow")
)

kable(error_summary, 
      caption = "Prediction Error Analysis - Environmental Patterns") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(c(4,5), background = "#ffe6e6")  # Highlight critical errors
```

```{r}
#| label: fake-critical-errors
#| echo: false

# Examples of critical prediction errors for learning
critical_errors_table <- data.frame(
  Date = c("2023-02-15", "2023-01-08", "2022-12-22", "2023-03-03", "2022-11-18"),
  Region = c("Glen Coe", "Lochaber", "N. Cairngorms", "Creag Meagaidh", "Torridon"),
  Actual = c(5, 1, 4, 2, 3),
  Predicted = c(2, 4, 1, 5, 1),
  Error = c(-3, +3, -3, +3, -2),
  Key_Conditions = c("Rapid warming + rain", "Surface hoar formation", 
                    "Wind slab development", "New snow + strong winds",
                    "Temperature inversion"),
  Lesson_Learned = c("Need better rain-on-snow detection", 
                    "Improve surface condition modeling",
                    "Enhanced wind loading algorithms",
                    "Better new snow weighting",
                    "Include elevation-specific temps")
)

kable(critical_errors_table, 
      caption = "Critical Prediction Errors - Case Studies for Model Improvement") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(c(6,7), width = "3cm")
```

## Model Deployment & Operational Considerations {#deployment}

```{r}
#| label: deployment-pipeline
#| eval: false
#| echo: true

# Production deployment pipeline for the avalanche prediction system
# This code outlines the operational implementation

# Model serving pipeline
library(plumber)
library(jsonlite)

# Create API endpoint for real-time predictions
#* @title Scottish Avalanche Hazard Prediction API
#* @description Predict avalanche hazard levels using neural network model

#* Predict avalanche hazard level
#* @param region:character Forecasting region (Lochaber, Glen Coe, etc.)
#* @param temp_min:numeric Minimum temperature in Celsius
#* @param temp_max:numeric Maximum temperature in Celsius  
#* @param wind_speed:numeric Wind speed in m/s
#* @param precipitation:numeric Precipitation in mm
#* @param snow_depth:numeric Snow depth in cm
#* @param elevation:numeric Elevation in meters
#* @post /predict
function(region, temp_min, temp_max, wind_speed, precipitation, snow_depth, elevation) {
  
  # Input validation
  required_regions <- c("Lochaber", "Glen Coe", "Creag Meagaidh", 
                       "Southern Cairngorms", "Northern Cairngorms", "Torridon")
  
  if (!region %in% required_regions) {
    return(list(error = "Invalid region specified"))
  }
  
  # Prepare input data
  input_data <- data.frame(
    region = region,
    temperature_min = as.numeric(temp_min),
    temperature_max = as.numeric(temp_max),
    wind_speed = as.numeric(wind_speed),
    precipitation = as.numeric(precipitation), 
    snow_depth = as.numeric(snow_depth),
    elevation = as.numeric(elevation)
  )
  
  # Apply preprocessing pipeline
  processed_input <- predict(preProcess_model, input_data)
  
  # Make prediction
  prediction_probs <- model %>% predict(as.matrix(processed_input))
  predicted_class <- which.max(prediction_probs) 
  confidence <- max(prediction_probs)
  
  # Format response
  response <- list(
    predicted_hazard_level = predicted_class,
    confidence = round(confidence, 3),
    probability_distribution = as.list(round(prediction_probs, 3)),
    timestamp = Sys.time(),
    model_version = "v2.1.0"
  )
  
  return(response)
}

# Model monitoring and drift detection
monitor_model_performance <- function(new_predictions, actual_outcomes) {
  
  # Calculate recent performance metrics
  recent_accuracy <- mean(new_predictions == actual_outcomes)
  
  # Compare against baseline performance
  baseline_accuracy <- 0.847  # From validation
  performance_decline <- baseline_accuracy - recent_accuracy
  
  # Alert if performance degrades significantly
  if (performance_decline > 0.05) {
    send_alert(paste("Model performance declined by", 
                    round(performance_decline * 100, 1), "%"))
  }
  
  # Check for data drift
  # Implementation would include statistical tests for feature distribution changes
}
```

```{r}
#| label: fake-deployment-metrics
#| echo: false

# Operational performance metrics
deployment_metrics <- data.frame(
  Metric = c("API Response Time", "Daily Predictions", "System Uptime", 
            "Model Accuracy (30-day)", "False Positive Rate", "False Negative Rate",
            "Critical Error Rate", "User Satisfaction", "Data Freshness"),
  Current_Value = c("127ms", "1,847", "99.7%", "0.834", "0.089", "0.074", 
                   "0.021", "4.2/5.0", "< 1 hour"),
  Target = c("< 200ms", "2,000+", "> 99.5%", "> 0.82", "< 0.10", "< 0.08",
           "< 0.03", "> 4.0/5.0", "< 2 hours"),
  Status = c("✅ Good", "✅ Good", "✅ Good", "✅ Good", "✅ Good", "✅ Good",
           "✅ Good", "✅ Good", "✅ Good")
)

kable(deployment_metrics, 
      caption = "Operational Performance Dashboard (Last 30 Days)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Conclusions & Future Work (This text is example!!!!) {#conclusions}

:::{.feature-grid}

:::{.card}
### 🎯 Key Findings
- **High Accuracy Achievement:** Neural network successfully predicts avalanche hazard with 84.7% accuracy across all hazard levels
- **Strong Feature Identification:** Snow depth, recent precipitation, and wind conditions emerge as most predictive variables
- **Regional Generalization:** Model performs consistently across all six Scottish forecasting regions (82.9% - 86.2% accuracy)
- **Seasonal Adaptability:** Best performance during peak winter months, acceptable performance in shoulder seasons
- **Critical Safety Focus:** Low false negative rate (7.4%) for high-risk conditions prioritizes mountain safety
:::

:::{.card}
### 🔮 Future Improvements
- **Real-time Data Integration:** Incorporate live weather station feeds and automated snow measurements
- **Ensemble Methods:** Combine neural networks with Random Forest and SVM models for robust predictions
- **Temporal Modeling:** Implement LSTM networks to capture time-series patterns in avalanche conditions
- **Uncertainty Quantification:** Add Bayesian neural networks for prediction confidence intervals
- **Multi-modal Data:** Integrate satellite imagery, webcam analysis, and crowdsourced observations
- **Edge Case Handling:** Improve performance on rare extreme weather events through synthetic data augmentation
:::

:::{.card}
### 🏔️ Practical Applications
- **SAIS Integration:** Enhanced decision support system for Scottish Avalanche Information Service forecasters
- **Mobile Application:** Real-time risk assessment app for backcountry users and mountain guides
- **Automated Alerts:** Early warning system for mountain rescue teams and ski patrol services
- **Educational Platform:** Training simulator for avalanche safety courses and professional development
- **Research Tool:** Foundation for advanced avalanche dynamics research and climate change impact studies
:::

:::


---

:::{style="background: linear-gradient(135deg, #667eea, #764ba2); color: white; padding: 2rem; border-radius: 15px; margin-top: 3rem; text-align: center;"}

### 🏔️ Project Impact Statement (This text is example!!!!)

*This research demonstrates the successful application of modern machine learning techniques to enhance avalanche hazard prediction in Scotland. By achieving 84.7% accuracy across six forecasting regions, the neural network model provides valuable decision support for mountain safety professionals while maintaining appropriate human oversight for life-critical applications.*

**The integration of AI tools throughout the development process showcases best practices for responsible AI adoption in safety-critical domains, emphasizing transparency, validation, and the complementary relationship between artificial intelligence and domain expertise.**

:::

---

*Final model deployment: March 2024 | Last updated: `r Sys.Date()` | Built with Quarto & R*